---
title: "Get example data"
output: html_document
bibliography: ../app.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

To demonstrate the package, we use DNA methylation data (450k array) deposited on GEO ([GSE42861](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE42861 "GSE42861")) [@liu_epigenome-wide_2013]. This dataset includes 450k data profiled from peripheral blood leukocyte (PBL) samples of 354 rheumatoid arthritis patiets and 335 controls. For this tutorial we'll use a subset of \~150 samples from this dataset.

# Download files

The files can be downloaded using the [GEOquery](https://www.bioconductor.org/packages/release/bioc/html/GEOquery.html "GEOquery") BioConductor R package.

```{r, eval = FALSE}
## Set directory
outdir <- "../GSE42861"

# Libraries 
library(dplyr)
library(magrittr)
library(stringr)

# Load samplesheet (using some very useful code included in the meffil package)
source("https://raw.githubusercontent.com/perishky/meffil/master/data-raw/geo.r")
samplesheet <- geo.samples("GSE42861")

# Format samplesheet 

## Get sample characteristics
dat <- stringr::str_split(samplesheet$characteristics_ch1, pattern=";", simplify=TRUE)

### Extract sample characteristics from `characteristics_ch1` column
characteristics <- list()
for(i in 1:ncol(dat)) {
  column_name <- stringr::str_trim(stringr::str_split(dat[,i], pattern=":", simplify=TRUE)[,1]) %>% unique()
  column_name <- stringr::str_replace(column_name, pattern=" ", replacement = "_")
  x <- stringr::str_trim(stringr::str_split(dat[,i], pattern=":", simplify=TRUE)[,2])
  tib <- tibble(x)
  colnames(tib) <- column_name
  characteristics[[i]] <- tib
}
characteristics <- dplyr::bind_cols(characteristics)

### Add 'Sample_Name' 
Sample_Name <- stringr::str_split(samplesheet$supplementary_file,
                                  pattern="/", simplify=TRUE)
Sample_Name <- Sample_Name[,ncol(Sample_Name)]
Sample_Name <- paste(stringr::str_split(Sample_Name, pattern="_", simplify=TRUE)[,1],
                     stringr::str_split(Sample_Name, pattern="_", simplify=TRUE)[,2],
                     stringr::str_split(Sample_Name, pattern="_", simplify=TRUE)[,3],
                     sep = "_")

### Combine 
samplesheet <- dplyr::bind_cols(tibble(Sample_Name = Sample_Name), 
                                characteristics, 
                                geo_accession = samplesheet$geo_accession)

### Add 'Basename' column -> path to IDAT files
samplesheet$Basename <- sprintf("%s/IDATs/%s", outdir, samplesheet$Sample_Name)

## Format column names 
samplesheet <- samplesheet %>%
  dplyr::mutate(
    Sex = toupper(gender),
    age = as.numeric(age),
    smoking_status = ifelse(smoking_status == "na", NA, smoking_status),
    Slide = stringr::str_split(Sample_Name,pattern="_",simplify=TRUE)[,2]
  ) %>%
  dplyr::rename(Smoking_Status = smoking_status, Age=age) %>%
  dplyr::select(-gender)

## Subset 
### For example purposes we'll use a subset of the data

slides_select <- c("7800246071", "5730053011", "573005038", "5730053041", "7800246085",
                   "7800246087", "7800246123", "7800246132", "7800246137",
                   
                   "5730192045", "5730053030", "5730053040", "5765205058", "5730192048", "7512560115", "5730053039")
samplesheet <- samplesheet %>%
  dplyr::filter(Slide %in% slides_select)

saveRDS(samplesheet, file=sprintf("%s/samplesheet_GSE42861.rds", outdir))

## Download IDAT files 
dir.create(sprintf("%s/IDATs", outdir), recursive=TRUE)
for(gsm in samplesheet$geo_accession) {
  getGEOSuppFiles(gsm, baseDir = sprintf("%s/IDATs", outdir), makeDirectory = FALSE)
}

# metrics <- get_qcmetrics(samplesheet)
# 
# QCapp(metrics, samplesheet=samplesheet %>% dplyr::mutate(Age=as.numeric(Age)))
```
